Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-07-29 supplement.tex

The number of publicly available whole-genome sequenced individuals from a
diverse set of African groups has grown in recent years
\citep[e.g.,][]{Mallick2016-lx,Bergstrom2020-cy,Schlebusch2020-zo}. Our choice
of populations was motivated by both technical and study design considerations.
First, while the diversity and linkage disequilibrium statistics used in our
inferences are robust to low coverage sequencing data \citep{Ragsdale2019-nt},
we wanted to have at least 20 individuals per population to increase accuracy
in statistical measurements by reducing the variance due to small sample sizes
\citep{Ragsdale2020-nz}. Second, merging of datasets post-variant calling can
introduce biases in demographic inference due to batch effects. We therefore
wanted to be able to perform joint calling in all of our samples. Together,
these considerations make the jointly called Thousand Genomes and ADRP dataset
best-suited for our demographic inference.

From a study design perspective, our choice of populations aimed to improve our
understanding of early human demography in Africa. The Nama derive ancestry
from a relatively deeply diverged population branch, and thus their demographic
history is of unique interest. Given previous archaeological and genetic
evidence, we expected that a minimal model for the Nama must include
pastoralist East African groups that contributed to the Nama
\citep[e.g.,][]{Uren2016-nn}, as well as recent European admixture, each of
which are represented in our jointly called dataset. Finally, because previous
studies have suggested a role for ``ghost'' archaic admixture or deep structure
in West African populations \citep[e.g.,][]{Speidel2019-nj,Durvasula2020-td},
and because these same sample sets are included in our dataset, we chose to
also include representatives of West African ancestry.

@@

\section{Model specification and fitting}
\label{modelspec}
For the early history, we tested model parameterizations that cover many of the
proposed scenarios of population structure, size changes, and/or archaic
hominin admixture. The simplest model, in terms of number of parameters, was a
single-origin expansion of humans, with no structure in the stem and no
archaic hominin admixture aside from the Neanderthal admixture in Eurasian populations
following the out-of-Africa migration. This model allowed for a population size
change in the stem of humans between the ancestral split of the
human-Neanderthal lineages and the more recent split of branches leading to
southern and western/eastern African populations.

@@

\subsection{Optimization using \moments}

@@

>>>>

The number of publicly available whole-genome sequenced individuals from a
diverse set of African groups has grown in recent years
\citep[e.g.,][]{Mallick2016-lx,Bergstrom2020-cy,Schlebusch2020-zo}. Our choice
of populations was motivated by both technical and study design considerations.
First, while the diversity and linkage disequilibrium statistics used in our
inferences are robust to low coverage sequencing data \citep{Ragsdale2019-nt},
we wanted to have at least 20 individuals per population to increase accuracy
in statistical measurements by reducing the variance due to small sample sizes
\citep{Ragsdale2020-nz}. Second, merging of datasets post-variant calling can
introduce biases in demographic inference due to batch effects (as demonstrated
in Section~\ref{sec:merged-bias}). We therefore wanted to be able to perform
joint calling in all of our samples. Together, these considerations make the
jointly called Thousand Genomes and ADRP dataset best-suited for our
demographic inference.

From a study design perspective, our choice of populations aimed to improve our
understanding of early human demography in Africa. The Nama derive ancestry
from a relatively deeply diverged population branch, and thus their demographic
history is of unique interest. Given previous archaeological and genetic
evidence, we expected that a minimal model for the Nama must include
pastoralist East African groups that contributed to the Nama
\citep[e.g.,][]{Uren2016-nn}, as well as European colonial admixture.  The
inclusion of three Ethiopian populations here represent some uncertainty in the
possible source population for the pastoralist contribution to the Nama, and
the Amhara and Oromo also have substantive ``back-to-Africa'' ancestry. The
Gumuz serve as a representative of southwest Ethipian ancestry that pre-dates
the Holocene back-to-Africa migration. Finally, because previous studies have
suggested a role for ``ghost'' archaic admixture or deep structure in West
African populations \citep[e.g.,][]{Speidel2019-nj,Durvasula2020-td}, we chose
to also include representatives of West African ancestry.

@@

\section{Model specification and fitting}
\label{modelspec}

For the early history, we tested model parameterizations that cover many of the
proposed scenarios of population structure, size changes, and/or archaic
hominin admixture. The simplest model, in terms of number of parameters, was a
single-origin expansion of humans, with no structure in the stem and no archaic
hominin admixture aside from the Neanderthal admixture in Eurasian populations
following the out-of-Africa migration. This model allowed for a population size
change in the stem of humans between the ancestral split of the
human-Neanderthal lineages and the more recent split of branches leading to
southern and western/eastern African populations. The recent expansion and the
African multi-regional models (Figure~\ref{fig:proposed-models}A and D) have
the same topology, so interpretation of the model depends on the specified or
inferred divergence times. Thus, because model optimization explores a wide
range of parameter values, we account for both scenarios with this
parameterization.

@@

\subsection{Optimization using \moments}
\label{sec:optimization}

@@

\subsection{Batch effects from merged data biases inference}\label{sec:merged-bias}

Merging samples from multiple independently called datasets can cause
systematic biases in computed diversity statistics (i.e., batch effects). In
order to explore the severity of batch effects on the LD statistics used in our
inferences, we took advantage of multiple independently sequenced and called
datasets using the same individuals from the Thousand Genomes project. We used
the phase 3 low coverage \citet{1000_Genomes_Project_Consortium2015-zq} data
and the high coverage resequenced data of the same individuals
\citep{Byrska-Bishop2021-jl} to 1) verify that computed diversity statistics
and inferences are robust to low coverage data, and 2) demonstrate that merging
data without joint variant calling can lead to substantial biases in inferred
demographic parameters. We analyzed two populations, the Mende (MSL) and
British (GBR), and considered a two-population isolation-with-migration model.

We computed four sets of diversity and LD statistics, one set of statistics
from the jointly called high coverage data, one from the jointly called low
coverage data, and one each from merged high and low coverage (MSL high
coverage and GBR low coverage; and GBR high coverage and MSL low coverage).
Using the measurement variances from each two-locus statistic in the high
coverage data, we estimated measurement error of the other three sets of
statistics (low/low, low/high, and high/low) against the high coverage data.
The jointly called low coverage data provided statistics that closely matched
the high coverage statistics, with each statistic within 1--2 standard errors.
On the other hand, both the merged datasets provided statistics that deviated
from the jointly called data, particularly for the cross-population statistics
(Figure~\ref{fig:supp-merged-data-SEs}.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=5in]{figures/supp-SE-merged-data}
    \caption{
        \textbf{Average error of statistics from low coverage and merged datasets.}
        Low coverage data provides consistent allelic diversity and linkage
        disequilibrium measures to high coverage data. On the other hand,
        merging separately called datasets can lead to large discrepancies
        in computed LD statistics.
    }
    \label{fig:supp-merged-data-SEs}
\end{figure}

With the observation that the low and high coverage data provided broadly
consistent statistics while the merged data provided inconsistent statistics,
we next wanted to assess whether their differences affect model inference. To
do this, we fit a two-population demographic model to the diversity and LD
statistics as described in Section~\ref{sec:optimization}. The modest
differences between the low and high coverage did not result in noticeable
differences in the optimized model (with an MSL-GBR inferred divergence time of
60ka, consistent with previous genomics studies). On the other hand, both
merged datasets provided substantially inconsistent fits, both to the high and
low coverage data and to each other. Notably, the inferred divergence time is
significantly older, due to the overestimation of mutational differences
between populations due to batch effects.

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/supp-merged-data-model-fits}
    \caption{
        \textbf{Merged datasets bias demographic inferences.}
        While statistics from low and high coverage datasets provide consistent
        best-fit demographic models, merged datasets result in biased LD
        statistics (Figure~\ref{fig:supp-merged-data-SEs}). These biased LD
        statistics lead to large downstream biases in inferred demographic
        models.
    }
    \label{fig:supp-merged-data-fits}
\end{figure}

More careful post-merging quality control may alleviate some of these
discrepancies. However, low coverage data provides a consistent model fit to
the high coverage data. Therefore, it is preferable to favor jointly called low
coverage data over a mix of high coverage but separately called data, at least
when using the diversity and LD statistics that we have in our analyses.
